Converging High-Throughput and High-Performance Computing: A Case Study

### 2017-04-03 overall.tex
Address backfill definition, Jack's tickets #9, #13, #14

Figure~\ref{fig:backfill-utilization} (gray bars) shows the number of core-hours
used by ATLAS during the experiment time window. ATLAS consumed a total of 39.3M
core-hours, for an average of $\approx$3.7M core-hours a month, with a minimum
of 1.7M core-hours in April 2016 and a maximum 7.8M core-hours in February 2017.

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/backfill_consumption.pdf}
    \caption{Gray bars: Number of unused cores of Titan's backfill availability;
    blue bars: Number of cores of Titan's backfill availability used by ATLAS
    via PanDA Brokers; red labels: Efficiency of PanDA Brokers defined as
    percentage of Titan's backfill availability used by ATLAS.}
\label{fig:backfill-utilization}
\end{figure}

Figure~\ref{fig:backfill-utilization} (blue bars) shows backfill utilization
during the experiment time window. Efficiency
(Figure~\ref{fig:backfill-utilization}, red labels) is defined as the fraction
of Titan core-hours utilized by the PanDA Brokers (for ATLAS jobs) to the total
of Titan’s backfill availability. The brokers reached 18\% average efficiency,
with a minimum 8.9\% efficiency on April 2016 and a maximum 33.5\% efficiency on
February 2017.
% The total backfill available was 38.1M Titan core-hours in April 2016, and
% 33.1M Titan core-hours in February 2017. This shows that the measured
% difference in efficiency did not depend on a comparable difference in total
% backfill availability.
The total backfill available was 20.3M Titan core-hours in April
2016, and 17.6M Titan core-hours in February 2017. This shows that the measured
difference in efficiency did not depend on a comparable difference in total
backfill availability.

% \mtnote{Elaborate further and put it at the end, after giving several of the
% measurements we would need. ``Currently, this measure of efficiency is being
% refined on the base of the actual minimum amount of walltime that PanDA Broker
% can request when submitting job requests. This is not a fixed value as it
% depends on variable initialization and configuration overheads, simulation time,
% and number of nodes obtained. Part of the undergoing effort is to enable the
% measurements of  the distribution of these parameters to derive an optimization
% function.''}

@@

During the experiment time window, about 2.25M detector simulation jobs were
completed on Titan, for a total of 225M events processed. This is equivalent to
0.9\% of all the 250M detector simulations performed by ATLAS in the same period
of time, and 3.5\% of the 6.6B events processed by those jobs.
% Comparatively, Titan contributed 3.9\% of the total of around
% 200K\mtnote{TODO: Fix this number} cores available to ATLAS on Grid, Cloud,
% and HPC infrastructures together.
These figures confirms the relevance of supercomputers' resource contribution to
the LHC Run 2, especially when accounting for the amount of unused backfill
availability and the improvement of PanDA efficiency across the experiment time
window.

@@

On February 2017, PanDA Brokers used almost twice as much backfill availability
than in any other month. No relevant code update was made during that period and
logs indicated that the brokers were able to respond more promptly to backfill
availability. This is likely due to hardware upgrades on the DTNs. The absence
of continuous monitoring of those nodes does not allow to quantify bottlenecks
but spot measurements of their load indicate that a faster CPU and better
networking were likely responsible for the improved performance.

Investigations showed an average CPU load of 3.6\% on the upgraded DTNs. As
such, further hardware upgrades seem unlikely to improve significantly the
performance of PanDA Brokers. Nonetheless, current DTN load suggests that the
number of brokers per DTN could be increased. In September 2016, increasing the
number of concurrent PanDA Broker from 4 to 20 markedly improved efficiency (see
Fig.~\ref{fig:backfill-utilization}) but further research is needed to
understand whether an even larger number brokers would allow PanDA to consume a
higher percentage of backfill availability.

Every detector simulation executed on Titan process 100 events. This number of
events is consistent with the physics of the use case and with the average
duration of backfill availability. The duration of a detector simulation is a
function of the number of events: the more events, the longer the simulation.
Not all events take the same time to be simulated:
% Figure~\ref{fig:100event-distrib} shows a distribution of the simulation time
% of
1 event takes from $\approx$2 to $\approx$40 minutes to be simulated, with a
mean of $\approx$14 minutes. Considering that each worker node process up to 16
events concurrently, 100 events takes around 2 hours to process. As such, PanDA brokers do not use backfill availability with less than 2 hours walltime.

@@

Usually, detector simulations performed on the Grid process 10 times the number
of events than those on Titan. This explains the difference between the
percentage of detector simulations (9\%) and of events computed by those
simulations (3.9\%) on Titan. PanDA Broker could fit the number of events to the
walltime of the backfill availability on the base of the distributions of
Fig.~\ref{fig:backfill-distrib}. That specific number of event could then be
pulled from the PanDA Event service~\cite{calafiura2015atlas} and given as input
to one or more simulations. Once packaged into the MPI script submitted to
titan's PBS batch system, these simulations would better fit backfill
availability, increasing the efficiency of PanDA Brokers.

@@

Currently, each PanDA Broker creates, submits, and monitor a single MPI PBS
script at a time. This design is inherited from PanDA Pilot where a single
process is spawn at a time to execute the payload. As a consequence, the
utilization of a larger portion of Titan's backfill availability depends on the
the number of concurrent PanDA Brokers instantiated on the DTNs: When all the 20
PanDA Brokers have submitted a MPI PBS Job, further backfill availability cannot
be used.

Based on the current data, increasing PanDA Brokers efficiency on Titan would
require around 60 concurrent brokers. Further testing and better monitoring
capabilities are required to establish whether the DTN capabilities can support
this load, especially when considering the volume of files staged in and out the
DTNs. Alternatively, a design of a PanDA Broker capable of managing multiple MPI
scripts at a time is being evaluated.

>>>>

We calculate the total amount of backfill availability over a period of time by:
(i) polling backfill availability at regular intervals during that time window;
(ii) converting the number of worker nodes available and their walltime into
core-hours; (iii) summing the number of core-hours. We call this number of
core-hours `total backfill availability'.

Figure~\ref{fig:backfill-utilization} shows the total backfill availability on
Titan (gray bars) and the number of core-hours of that availability used by
ATLAS (blue bars) during the experiment time window. ATLAS consumed a total of
51.4M core-hours, for an average of $\approx$3.7M core-hours a month, with a
minimum of 1.7M core-hours in April 2016 and a maximum 7.9M core-hours in
February 2017.

\begin{figure}[htp]
    \includegraphics[clip,width=\columnwidth]{figures/backfill_consumption.pdf}

    \caption{Titan's total backfill availability: CPU core-hours (gray) and CPU
    core-hours used by ATLAS (blue). GPU core-hours unaccounted for as they
    cannot be used by ATLAS. Efficiency of PanDA Brokers defined as percentage
    of total Titan's backfill availability used by ATLAS (Red labels).}

\label{fig:backfill-utilization}
\end{figure}

PanDA Brokers' efficiency (Figure~\ref{fig:backfill-utilization}, red labels) is
defined as the fraction of core-hours utilized by the PanDA Brokers to  Titan’s
total backfill availability during the experiment time window. The brokers
reached 18\% average efficiency, with a minimum 7.8\% efficiency on May 2016 and
a maximum 30.9\% efficiency on February 2017.
% The total backfill available was 38.1M Titan core-hours in April 2016, and
% 33.1M Titan core-hours in February 2017. This shows that the measured
% difference in efficiency did not depend on a comparable difference in total
% backfill availability.
The total backfill availability was 20.3M in April 2016, and 17.6M in February
2017. This shows that the measured difference in efficiency did not depend on a
comparable difference in the total backfill availability.

% \mtnote{Elaborate further and put it at the end, after giving several of the
% measurements we would need. ``Currently, this measure of efficiency is being
% refined on the base of the actual minimum amount of walltime that PanDA Broker
% can request when submitting job requests. This is not a fixed value as it
% depends on variable initialization and configuration overheads, simulation
% time, and number of nodes obtained. Part of the undergoing effort is to
% enable the measurements of  the distribution of these parameters to derive an
% optimization function.''}

@@

During the experiment time window, about 2.25M detector simulation jobs were
completed on Titan, for a total of 225M events processed. This is equivalent to
0.9\% of all the 250M detector simulations performed by ATLAS in the same period
of time, and 3.5\% of the 6.6B events processed by those jobs.
% Comparatively, Titan contributed 3.9\% of the total of around
% 200K\mtnote{TODO: Fix this number} cores available to ATLAS on Grid, Cloud,
% and HPC infrastructures together.
These figures confirms the relevance of supercomputers' resource contribution to
the LHC Run 2, especially when accounting for the amount of unused total
backfill availability and the improvement of PanDA efficiency across the
experiment time window.

@@

On February 2017, PanDA Brokers used almost twice as much total backfill
availability than in any other month (preliminary results for March 2017
displayed in Fig.~\ref{fig:backfill-utilization} confirm this trend). No
relevant code update was made during that period and logs indicated that the
brokers were able to perform faster. This is likely due to hardware upgrades on
the DTNs. The absence of continuous monitoring of those nodes does not allow to
quantify bottlenecks but spot measurements of their load indicate that a faster
CPU and better networking were likely responsible for the improved performance.
Investigations showed an average CPU load of 3.6\% on the upgraded DTNs. As
such, further hardware upgrades seem unlikely to improve significantly the
performance of PanDA Brokers.

Every detector simulation executed on Titan process 100 events. This number of
events is consistent with the physics of the use case and with the average
duration of backfill availability. The duration of a detector simulation is a
function of the number of events simulated but not all events take the same time
to be simulated.
% Figure~\ref{fig:100event-distrib} shows a distribution of the simulation time
% of
1 event simulation takes from $\approx$2 to $\approx$40 minutes, with a mean of
$\approx$14 minutes. Considering that each worker node process up to 16 events
concurrently, 100 events takes an average of 105 minutes to process. As such,
PanDA brokers do not use backfill availability with less than 105 minutes
walltime.

@@

Usually, detector simulations performed on the Grid process 10 times the number
of events than those on Titan. This explains the difference between the
percentage of detector simulations and of events computed by those simulations
on Titan. PanDA Broker could fit the number of events to the walltime of the
backfill availability on the base of the distributions of
Fig.~\ref{fig:backfill-distrib}. That specific number of event could then be
pulled from the PanDA Event service~\cite{calafiura2015atlas} and given as input
to one or more simulations. Once packaged into the MPI script submitted to
titan's PBS batch system, these simulations would better fit backfill
availability, increasing the efficiency of PanDA Brokers.

@@

Currently, each PanDA Broker creates, submits, and monitor a single MPI PBS
script at a time. This design is inherited from PanDA Pilot where a single
process is spawn at a time to execute the payload. As a consequence, the
utilization of a larger portion of Titan's total backfill availability depends
on the the number of concurrent PanDA Brokers instantiated on the DTNs: When all
the 20 PanDA Brokers have submitted a MPI PBS Job, further backfill availability
cannot be used.

In September 2016, increasing the number of concurrent PanDA Broker from 4 to 20
markedly improved efficiency (see Fig.~\ref{fig:backfill-utilization}) but
further research is undergoing to understand whether an even larger number
brokers would yield similar results. This research focuses on evaluating the the
overheads of input/output files staging, including its impact on DTNs, and on an
alternative design of PanDA Broker to enable the concurrent submission of
multiple MPI scripts~\cite{barreiro2016panda}.
